---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# rctbayespower

<!-- badges: start -->
<!-- Note: Badges are placeholders for when the package is published -->
<!-- badges: end -->

## Disclaimer

This is an experimental package written with the help of LLMs and has not been 
thoroughly tested yet.
Hence, I do not recommend using it for production code or critical analyses.



## Installation

You can install the development version of `rctbayespower` from GitHub:

```{r, eval = FALSE}
# install.packages("devtools")
devtools::install_github("matthiaskloft/rctbayespower")
```



## Overview

`rctbayespower` provides tools for conducting Bayesian power analysis for randomized controlled trials (RCTs) using `brms` and Stan. The package allows researchers to:

### Multiple Outcome Types

- **Continuous outcomes**: Using Cohen's d effect sizes
- **Binary outcomes**: Using log odds ratios
- **Count outcomes**: Using log rate ratios


### Flexible Power Metrics

- **Probability of Success/Futility**: Probability effect exceeds success/futility thresholds
- **Significance Power**: Traditional frequentist-like power


### Advanced Features

- Custom prior specifications
- Power curves and sample size curves
- Comprehensive plotting functions



## Ressources

The best way to get started is to visit the package website and the introductory vignette:

- Website: [https://matthiaskloft.github.io/rctbayespower/](https://matthiaskloft.github.io/rctbayespower/)

- Vignettes:
  - [Introduction](https://matthiaskloft.github.io/rctbayespower/articles/01-introduction.html): Basic usage and concepts
  - [Case Studies](https://matthiaskloft.github.io/rctbayespower/articles/02-case-studies.html): Real-world applications
  - [Prior Specifications](https://matthiaskloft.github.io/rctbayespower/articles/03-prior-specifications.html): Working with priors
  - [Advanced Modeling](https://matthiaskloft.github.io/rctbayespower/articles/04-advanced-modeling.html): Complex designs
  - [Algorithm Performance](https://matthiaskloft.github.io/rctbayespower/articles/05-algorithm-performance.html): Performance optimization

- [Open an Issue](https://github.com/matthiaskloft/rctbayespower/issues)


