---
title: "Getting Started with 'rctbayespower'"
author: 
 - name: Matthias Kloft
   orcid: 0000-0003-1845-6957
date: "`r Sys.Date()`"
format:
  html:
    toc: true
    number-sections: true
    theme: cosmo
    code-fold: false
    code-tools: true
    code-summary: "Show the code"
    fig-width: 7
    fig-height: 4.5
    embed-resources: true
execute:
  message: false
  warning: false
---

Setup:
```{r setup}
#| message: false

# Set CRAN mirror
options(repos = c(CRAN = "https://cloud.r-project.org/"))

# Check and install required packages
packages <- c("dplyr", "kableExtra", "pander")
install.packages(setdiff(packages, rownames(installed.packages())))
invisible(lapply(packages, require, character.only = TRUE))
library(rctbayespower)


# Set number of cores and simulations:
n_cores <- parallel::detectCores() - 2
n_sims <- 500
```

***

# Introduction

The `rctbayespower` package provides tools for conducting Bayesian power analysis for randomized controlled trials (RCTs) using the `brms` package and Stan. 



# Basic Power Analysis Using ANCOVA Design



## Model

Show predefined models available in the package:
```{r}
list_predefined_models()
```


Get the predefined ANCOVA model for continuous outcomes with two arms:
```{r build_model_ancova}
model_ancova <- build_model(predefined_model = "ancova_cont_2arms")
print(model_ancova)
```

## Design

Find possible target parameters in the model.
```{r}
model_ancova@parameter_names_brms
```

Specify the study design with target parameter, thresholds for success and futility, and significance levels for frequentist-like power.
```{r build_design}
design <- build_design(
  model = model_ancova,
  target_params = "b_arm2",
  p_sig_scs = 0.9,
  p_sig_ftl = 0.5,
  analysis_at = c(.3,.6,1),
  adaptive = FALSE
)
print(design)
```


### Conditions

Find parameters that need user specification for the design.
```{r}
required_fn_args(design)
```

Specify conditions.
```{r build_conditions}
conditions <- build_conditions(
  design = design,
  condition_values = list(
    # two sample sizes
    n_total = seq(80, 160, by = 80),
    # two effect sizes
    b_arm_treat = c(0.3)
  ),
  static_values = list(
    # baseline effect
    b_covariate = 0,
      thresholds_success = 0.1,
  thresholds_futility = 0
  )
)
print(conditions, n = 100)
```


## Run Power Analysis

Run the analysis using the `power_analysis` function. This will run the model for each condition and return the results.
```{r run_sim}
power <- power_analysis(
  conditions = conditions,
  n_cores = n_cores,
  n_sims = n_cores,
  verbosity = 2,
  brms_args = list(
    chains = 1,
    iter = 300,
    warmup = 100
  )
)
```

```{r}
print(power)
```

```{r}
plot(power, type = "power_curve", facet_by = "effect_size")
```


## Power for Effect of 0.3
```{r power_success}
power@results_summ |>
  dplyr::filter(b_arm_treat != 0) |>
  select(
    n_total,
    b_arm_treat,
    pr_scs,
    se_pr_scs,
    pwr_scs,
    se_pwr_scs,
    pr_ftl,
    se_pr_ftl,
    pwr_ftl,
    se_pwr_ftl
  ) |>
  dplyr::arrange(desc(b_arm_treat)) |>
  kableExtra::kable(digits = 3, format = "html")
```


## Alpha Error Rate for Null Effect 

The power for success indicates the alpha error rate for the null effect (b_arm_treat = 0). This is the probability of rejecting the null hypothesis when it is true, which should be close to the significance level (0.05) if the design is well specified.

```{r power_null}
power@results_summ |>
  dplyr::filter(b_arm_treat == 0) |>
  select(
    n_total,
    b_arm_treat,
    pr_scs,
    se_pr_scs,
    pwr_scs,
    se_pwr_scs,
    pwr_ftl,
    se_pwr_ftl,
    pr_ftl, 
    prob_futility_se 
  ) |>
  dplyr::arrange(desc(b_arm_treat)) |> 
  kableExtra::kable(digits = 3, 
                    format = "html")
```

## Run Time

```{r}
#| code-fold: true


n_runs <- nrow(power@conditions@conditions_grid) * power@n_sims
cat("Total run time:", round(power@elapsed_time,1), "minutes for", n_runs, "total simulation repetitions using", power@n_cores, "cores.\n")
```

***


# Session Information

```{r session_info}
pander::pander(sessionInfo())
```
