---
title: "Prior Specifications in 'rctbayespower'"
author: 
 - name: Matthias Kloft
   orcid: 0000-0003-1845-6957
date: "`r Sys.Date()`"
format:
  html:
    toc: true
    number-sections: true
    theme: cosmo
    code-fold: true
    code-tools: true
    code-summary: "Show the code"
    fig-width: 7
    fig-height: 4.5
    embed-resources: true
execute:
  message: false
  warning: false
---

Setup:
```{r setup}
# Set CRAN mirror
options(repos = c(CRAN = "https://cloud.r-project.org/"))

# Check and install required packages
packages <- c("dplyr", "kableExtra", "pander")
install.packages(setdiff(packages, rownames(installed.packages())))
invisible(lapply(packages, require, character.only = TRUE))
library(rctbayespower)

# Set number of cores and simulations:
n_cores <- parallel::detectCores() - 2
n_sims <- 500
```


***


The `rctbayespower` package uses two types of priors: **estimation priors** (for Bayesian model fitting) and **design priors** (for integrating effect size uncertainty). This vignette demonstrates their impact on power analysis through two focused examples.

# Example 1: Different Estimation Priors

Estimation priors affect how treatment effects are estimated during Bayesian model fitting. Here we compare default priors versus informative priors.


Ancova model with two arms is used to illustrate the impact of different priors on power analysis results. We will compare uninformative and sceptical priors.

Let's first define the sample size for our analysis:
```{r}
sample_size <- 100
```

## Analysis of Medium Effect of Cohen's d = 0.5





### Analysis with uninformative priors

Uninformative, locally flat prior on treatment effect:
$$d \sim \mathcal{N}(0,100)$$


```{r}
# Create ANCOVA model with uninformative priors
uninformative_model <- build_model_ancova_cont_2arms(
  prior_treatment = brms::set_prior("normal(0, 100)", class = "b"),
  prior_covariate = brms::set_prior("normal(0, 100)", class = "b", coef = "covariate"),
  prior_intercept = brms::set_prior("normal(0, 100)", class = "Intercept"),
  prior_sigma = brms::set_prior("normal(0, 100)", class = "sigma", lb = 0),
  link_sigma = "identity"
)

# Create design from custom model
uninformative_design <- build_design(
  sim_fn = uninformative_model@sim_fn,
  inference_model = uninformative_model@inference_model,
  n_endpoints = 1,
  endpoint_types = "continuous",
  n_arms = 2,
  target_params = "b_arm2"
)

# Create conditions
uninformative_conditions <- build_conditions(
  design = uninformative_design,
  crossed = list(n_total = sample_size),
  constant = list(
    b_arm_treat = 0.5,        # Medium effect
    b_covariate = 0,          # No covariate effect
    p_alloc = list(c(0.5, 0.5)),
    intercept = 0,
    sigma = 1,
    thr_dec_eff = 0.95,       # Probability threshold for efficacy
    thr_dec_fut = 0.5,        # Probability threshold for futility
    thr_fx_eff = 0.3,         # Effect size threshold for efficacy
    thr_fx_fut = 0.1          # Effect size threshold for futility
  )
)

# Run power analysis
uninformative_analysis <- rctbayespower::power_analysis(
  conditions = uninformative_conditions,
  n_sims = n_sims,
  n_cores = n_cores,
  verbosity = 0
)
```

```{r}
res_uninformative <- uninformative_analysis@results_conditions |>
  select(
    n_total,
    post_med,
    post_sd,
    pr_eff,
    pwr_eff,
    pr_fut,
    pwr_fut
  ) |>
  dplyr::mutate(Prior = "uninformative", .before = everything())

res_uninformative |> 
  kableExtra::kable(digits = 3, format = "html")
```


Analysis with Informative Prior

Informative, sceptical prior on treatment effect:

$$d \sim \mathcal{t_3}(0,1)$$

```{r}
# Create ANCOVA model with informative/sceptical priors
informative_model <- build_model_ancova_cont_2arms(
  prior_treatment = brms::set_prior("student_t(3, 0, 2)", class = "b"),
  prior_covariate = brms::set_prior("student_t(3, 0, 2)", class = "b", coef = "covariate"),
  prior_intercept = brms::set_prior("student_t(3, 0, 2)", class = "Intercept"),
  prior_sigma = brms::set_prior("student_t(3, 0, 2)", class = "sigma", lb = 0),
  link_sigma = "identity"
)

# Create design from custom model
informative_design <- build_design(
  sim_fn = informative_model@sim_fn,
  inference_model = informative_model@inference_model,
  n_endpoints = 1,
  endpoint_types = "continuous",
  n_arms = 2,
  target_params = "b_arm2"
)

# Create conditions
informative_conditions <- build_conditions(
  design = informative_design,
  crossed = list(n_total = sample_size),
  constant = list(
    b_arm_treat = 0.5,        # Medium effect
    b_covariate = 0,          # No covariate effect
    p_alloc = list(c(0.5, 0.5)),
    intercept = 0,
    sigma = 1,
    thr_dec_eff = 0.975,      # Probability threshold for efficacy
    thr_dec_fut = 0.5,        # Probability threshold for futility
    thr_fx_eff = 0.3,         # Effect size threshold for efficacy
    thr_fx_fut = 0.1          # Effect size threshold for futility
  )
)

# Run power analysis
informative_analysis <- power_analysis(
  conditions = informative_conditions,
  n_sims = n_sims,
  n_cores = n_cores,
  verbosity = 0
)
```

```{r}
res_informative <- informative_analysis@results_conditions |>
  select(
    n_total,
    post_med,
    post_sd,
    pr_eff,
    pwr_eff,
    pr_fut,
    pwr_fut
  ) |>
  dplyr::mutate(Prior = "informative", .before = everything())

res_informative |> 
  kableExtra::kable(digits = 3, format = "html")
```



Comparing the results of the two analyses:
```{r}
estimation_comparison <- 
  rbind(
    res_uninformative,
    res_informative
  ) 

estimation_comparison|>
  # round the numeric columns for better readability
  kableExtra::kable(digits = 3)
```


## Alpha Error Rate

To evaluate the alpha error rate (false positive rate) for both prior configurations, we need to run the power analysis under the null hypothesis (effect size = 0) and examine how often we incorrectly conclude for success. This tells us how well our Bayesian decision framework controls Type I error.
Luckily, the `rctbayespower` package allows us to reuse the same model and design configurations, only changing the effect size to zero.

```{r alpha_error_uninformative}
# Reuse uninformative design (same model)
# Create conditions with null hypothesis (no effect)
alpha_uninformative_conditions <- build_conditions(
  design = uninformative_design,
  crossed = list(n_total = sample_size),
  constant = list(
    b_arm_treat = 0,          # Null hypothesis: no effect
    b_covariate = 0,          # No covariate effect
    p_alloc = list(c(0.5, 0.5)),
    intercept = 0,
    sigma = 1,
    thr_dec_eff = 0.95,       # Probability threshold for efficacy
    thr_dec_fut = 0.5,        # Probability threshold for futility
    thr_fx_eff = 0,           # Effect size threshold for efficacy (null)
    thr_fx_fut = 0            # Effect size threshold for futility (null)
  )
)

alpha_uninformative <- power_analysis(
  conditions = alpha_uninformative_conditions,
  n_sims = n_sims,
  n_cores = n_cores,
  verbosity = 0
)
```

```{r alpha_error_informative}
# Reuse informative design (same model)
# Create conditions with null hypothesis (no effect)
alpha_informative_conditions <- build_conditions(
  design = informative_design,
  crossed = list(n_total = sample_size),
  constant = list(
    b_arm_treat = 0,          # Null hypothesis: no effect
    b_covariate = 0,          # No covariate effect
    p_alloc = list(c(0.5, 0.5)),
    intercept = 0,
    sigma = 1,
    thr_dec_eff = 0.95,       # Probability threshold for efficacy
    thr_dec_fut = 0.5,        # Probability threshold for futility
    thr_fx_eff = 0,           # Effect size threshold for efficacy (null)
    thr_fx_fut = 0            # Effect size threshold for futility (null)
  )
)

alpha_informative <- power_analysis(
  conditions = alpha_informative_conditions,
  n_sims = n_sims,
  n_cores = n_cores,
  verbosity = 2
)
```

Comparing the alpha error rates:
```{r alpha_comparison}
res_alpha <- 
  rbind(
    alpha_uninformative@results_conditions |>
      select(
        n_total,
        pr_eff,
        pwr_eff
      ) |>
      dplyr::mutate(Prior = "uninformative", .before = everything()),
    alpha_informative@results_conditions |>
      select(
        n_total,
        pr_eff,
        pwr_eff
      ) |>
      dplyr::mutate(Prior = "informative", .before = everything())
  ) |> 
  dplyr::rename(
    
    alpha = pwr_eff,
    "p(d > 0)" = pr_eff,
  )

res_alpha |> 
  kableExtra::kable(digits = 3, format = "html")
```

Using a sceptical prior reduces the probability of falsely concluding success when there is no true effect, thus controlling the alpha error rate more effectively than the uninformative priors.


# Example 2: Design Priors

Coming soon ...


***


```{r session_info}
pander::pander(sessionInfo())
```
