---
title: "Backend Comparison"
subtitle: "brms vs. BayesFlow: Multicore CPU vs. GPU Accelerated Performance"
author:
 - name: Matthias Kloft
   orcid: 0000-0003-1845-6957
date: "`r Sys.Date()`"
format:
  html:
    toc: true
    number-sections: true
    theme: cosmo
    code-fold: false
    code-tools: true
    code-summary: "Show the code"
    fig-width: 7
    fig-height: 4.5
    embed-resources: true
execute:
  message: true
  warning: true
---

```{r setup}
#| message: false

library(rctbayespower)
library(tidyverse)
library(pander)

# install venv
if(!reticulate::virtualenv_exists("rctbp-3-12")){
  setup_bf_python(envname = "rctbp-3-12", python_version = "3.12", cuda_version = "auto")
}
# Initialize BayesFlow Python environment
init_bf("rctbp-3-12")

n_sims <- 1000L
n_cores <- parallel::detectCores() - 1
```

***

# Introduction

Power analysis for Bayesian trials typically requires running thousands of simulations, each involving full posterior estimation. With traditional MCMC methods, large simulation studies can take hours. The **BayesFlow backend** reduces this to minutes.

## Two Backends

The `rctbayespower` package supports two backends for posterior estimation:

| | brms/Stan | BayesFlow |
|--------|-----------|-----------|
| **Method** | MCMC sampling | Neural posterior estimation |
| **GPU support** | No | Yes |
| **CPU parallelization** | Yes | No |
| **Setup** | R only | Requires Python |

**BayesFlow** uses neural networks trained on simulated data to learn the mapping from summary statistics to posterior distributions. Once trained, inference is a single forward pass—nearly instantaneous.

## When to Use Each

- **brms/Stan**: Final analyses, validation, complex custom models
- **BayesFlow**: Rapid prototyping, parameter sweeps, large simulation studies, Bayesian optimization


## Setting Up the Python Environment

The BayesFlow backend requires Python with PyTorch, Keras, and BayesFlow. Use `setup_bf_python()` for automated setup with GPU detection:

```{r setup-python}
#| eval: false

# One-command setup (first time only)
setup_bf_python()

# With options:
setup_bf_python(envname = "rctbp-3-12-gpu", python_version = "3.12", cuda_version = "auto")
setup_bf_python(envname = "rctbp-3-12-cpu", python_version = "3.12", cuda_version = "cpu")
setup_bf_python(cuda_version = "cpu")
setup_bf_python(force = TRUE)
```

Check your installation:

```{r bf-status}
#| eval: false

bf_status()
```

```
── BayesFlow Environment Status ──
── Python Configuration ──
ℹ Python: ~/.virtualenvs/r-rctbayespower/Scripts/python.exe
ℹ Version: 3.12
ℹ Virtual env: ~/.virtualenvs/r-rctbayespower

── Package Status ──
✔ Python: 3.12
✔ PyTorch: 2.9.1+cu128
✔ CUDA: 12.8 (NVIDIA GeForce GTX 1660 SUPER)
✔ BayesFlow: 2.0.7
✔ Keras: 3.12.0
✔ NumPy: 1.26.4

── Status ──
✔ BayesFlow ready with GPU acceleration
```

::: {.callout-note collapse="true"}
## Manual Installation

```bash
# Create and activate virtual environment
python -m venv ~/.virtualenvs/r-rctbayespower
source ~/.virtualenvs/r-rctbayespower/bin/activate  # Linux/Mac
# or: ~/.virtualenvs/r-rctbayespower/Scripts/activate  # Windows

# Install PyTorch with CUDA
pip install torch torchvision torchaudio --index-url https://download.pytorch.org/whl/cu128

# Install dependencies
pip install bayesflow keras numpy
```

Then in R: `reticulate::use_virtualenv("r-rctbayespower", required = TRUE)`
:::

***

# BayesFlow Workflow

The workflow with BayesFlow is nearly identical to brms.

## Step 1: Build Design

```{r build-design}
design <- build_design(
  model_name = "ancova_cont_2arms",
  backend = "bf",
  target_params = "b_group"
  )
print(design)
```

## Step 2: Build Conditions

```{r build-conditions}
conditions <- build_conditions(
  design = design,
  crossed = list(
    n_total = c(60, 80, 100, 120),
    b_arm_treat = c(0.5)
  ),
  constant = list(
    b_covariate = 0.3,
    thr_fx_eff = 0.1,
    thr_fx_fut = 0, 
    thr_dec_eff = 0.975,
    thr_dec_fut = 0.5
  )
)
print(conditions)
```

## Step 3: Run Power Analysis

```{r power-analysis-bf}
power_bf <- power_analysis(
  conditions = conditions,
  n_sims = n_sims,
  n_cores = 1,
  run = TRUE,
  verbosity = 2,
  bf_args = list(n_posterior_samples = 1000)
)
```

```{r power-results-bf}
print(power_bf)

power_bf@results_conditions |>
  select(id_cond, n_total, pwr_eff, pwr_fut, post_mn)
```

***

# Comparison with brms Backend

## Run Power Analysis with brms

```{r power-analysis-brms}
design_brms <- build_design(
  model_name = "ancova_cont_2arms", 
  backend = "brms",
  target_params = "b_arm2"
)
conditions_brms <- conditions
conditions_brms@design <- design_brms

power_brms <- power_analysis(
  conditions_brms,
  n_sims = n_sims,
  n_cores = n_cores,
  run = TRUE,
  verbosity = 0,
  brms_args = list(
    chains = 4,
    iter = 350,
    warmup = 100
  )
)
print(power_brms)
```

## Compare Timings

```{r timings}
bf_time <- power_bf@elapsed_time * 60
brms_time <- power_brms@elapsed_time * 60

format_time <- function(time_sec) {
  mins <- floor(time_sec / 60)
  secs <- round(time_sec %% 60)
  paste0(mins, " min ", secs, " sec")
}

speedup <- brms_time / bf_time

cat("BayesFlow:", format_time(bf_time), "\n")
cat("brms:", format_time(brms_time), "\n")
cat("Speedup:", round(speedup, 1), "x\n")
```

## Compare Power Curves
```{r}
results_bf <- power_bf@results_conditions %>%
  mutate(Backend = "BayesFlow")
results_brms <- power_brms@results_conditions %>%
  mutate(Backend = "brms")
combined_results <- rbind(results_bf, results_brms) |>
  pivot_longer(
    cols = c(pwr_eff, pr_eff),
    names_to = "Criterion",
    values_to = "Value"
  )
ggplot(combined_results, aes(x = n_total, y = Value, color = Backend)) +
  geom_line() +
  geom_point() +
  scale_y_continuous(limits = c(0, 1), breaks = seq(0, 1, by = 0.1)) +
  facet_wrap(~ Criterion) +
  labs(title = "Power Curves: BayesFlow vs brms", x = "Total Sample Size", y = "Power (Success Criterion)") +
  theme_minimal()
```


***

# Session Information

```{r session-info}
pander::pander(sessionInfo())
```
