---
title: "Bayesian Optimization for Trial Design"
author:
 - name: Matthias Kloft
   orcid: 0000-0003-1845-6957
date: "`r Sys.Date()`"
format:
  html:
    toc: true
    number-sections: true
    theme: cosmo
    code-fold: false
    code-tools: true
    code-summary: "Show the code"
    fig-width: 7
    fig-height: 5
    embed-resources: true
execute:
  message: true
  warning: true
---

```{r setup}
#| message: false

library(rctbayespower)
library(tidyverse)
library(reticulate)

backend <- "bf"

if (backend == "bf") {
  # install venv
  if (!reticulate::virtualenv_exists("rctbp-3-12")) {
    setup_bf_python(
      envname = "rctbp-3-12",
      python_version = "3.12",
      cuda_version = "auto",
      force = TRUE
    )
  }
  # Initialize BayesFlow Python environment
  init_bf("rctbp-3-12")
}

rctbayespower::check_bf_available()
```

***

# Introduction

Finding optimal trial designs often requires exploring a large parameter space. Traditional grid search evaluates every combination, which becomes computationally prohibitive as dimensions grow.

**Bayesian Optimization (BO)** efficiently searches continuous parameter spaces by:

1. Building a surrogate model (Gaussian Process) of the objective function
2. Using an acquisition function to select promising points to evaluate
3. Iteratively refining the search based on observed results

This is ideal for power analysis where each evaluation (running simulations) is expensive.

## Grid Search vs. Bayesian Optimization

| | Grid Search | Bayesian Optimization |
|--------|-----------|-----------|
| **API** | `build_conditions()` → `power_analysis()` | `build_objectives()` → `optimization()` |
| **Search** | Discrete grid points | Continuous bounds |
| **Evaluations** | All points | Guided selection |
| **Use case** | Visualization, known ranges | Sample size determination |

***

# Workflow

## Step 1: Create Design

Use the BayesFlow backend for faster evaluations.
```{r create-design}
design <- build_design(
 model_name = "ancova_cont_2arms",
 backend = backend,
 target_params = "b_group"
)
```

## Step 2: Define Optimization Problem

Use `build_objectives()` to specify:

- **search**: Parameter bounds to optimize over
- **objectives**: What to achieve (target power, minimize sample size)
- **constant**: Fixed design parameters

```{r}
show_objective_fns()
#obj_min_n_for_power()
```


```{r build-objectives}
obj <- build_objectives(
 design = design,
 search = list(n_total = c(50, 800)),
 objectives = list(pwr_eff = target(0.9, direction = ">=")),
 constant = list(
   p_alloc = .5,
   b_arm_treat = 0.3,
   thr_dec_eff = 0.95,
   thr_dec_fut = 0.5,
   thr_fx_eff = 0.1,
   thr_fx_fut = 0,
   intercept = 0,
   b_covariate = 0.3,
   sigma = 1
 )
)
#print(obj)
```

## Step 3: Run Optimization


### Configuration Options

```{r show-args}
#show_optimization_args()
```

### Run Optimization
```{r run-optimization, results='hide'}
init_bf("rctbp-3-12")
result <- optimization(
  obj,
  n_sims = c(100, 200, 500),
  sims_final_run = 1000,
  evals_per_step = c(20, 20, 10),
  init_design_size = 20,
  trim_param_space = .25,
  max_final_evals = 1,
  acq_function  = "cb",
  penalty = "logit_distance",
  penalty_kappa = 2,
  penalty_alpha = 0.5,
  surrogate = "gp",
  verbosity = 1,
  refresh = 10,
  bf_args = list(n_posterior_samples = 1000)
)
```

```{r}
library(data.table)

# Or create a fine grid for plotting:
new_data <- data.table(
  n_total = seq((50), (600), 1)
)

# After surrogate is fitted and updated:
result@mbo_objects$surrogate$update()

# Predict returns a data.table with 'mean' and 'se' columns
predictions <- result@mbo_objects$surrogate$predict(new_data)

# Access results:
plogis(predictions$mean)  # Vector of posterior means
plogis(predictions$se)    # Vector of posterior standard errors

# Combine for analysis/plotting:
new_data[, `:=`(
  mean = plogis(predictions$mean),
  se = plogis(predictions$se),
  n_total = ceiling((n_total))  # Back-transform to original scale
)]
new_data

# plot profile
new_data |> 
  as.data.frame() |> 
ggplot(aes(x = n_total, y = mean)) +
  geom_line() +
  geom_ribbon(aes(
    ymin = mean - 1.96 * se,
    ymax = mean + 1.96 * se
  ), alpha = 0.2) +
  geom_hline(yintercept = 0.9,
             linetype = "dashed",
             color = "grey20") +
  geom_vline(xintercept = result@result$n_total,
             linetype = 1,
             color = "red") +  
  labs(title = "Surrogate Model Prediction", x = "Total Sample Size", y = "Power for Efficacy") +
  scale_y_continuous(limits = c(0, 1)) +
  theme_minimal()
```

### Plot sampled points: n_total vs. time

```{r}
result@archive |>
  mutate(id_draw = row_number()) |>
  ggplot(aes(x = id_draw, y = pwr_eff)) +
  geom_hline(yintercept = 0.9,
             linetype = "dashed",
             color = "grey20") +
  geom_hline(yintercept = result@result$pwr_eff,
             linetype = 1,
             color = "red") +           
  geom_point() +
  ggplot2::geom_errorbar(aes(
    ymin = pwr_eff - 1.96 * se_pwr_eff,
    ymax = pwr_eff + 1.96 * se_pwr_eff
  )) +
  geom_line() +
  scale_y_continuous(limits = c(0, 1)) +
  labs(title = "Sampled Points during Optimization", x = "Draw Number", y = "Power for Efficacy") +
  theme_minimal()

result@archive |>
  mutate(id_draw = row_number()) |>
  ggplot(aes(x = n_total, y = pwr_eff)) +
  geom_hline(yintercept = 0.9,
             linetype = "dashed",
             color = "grey20") +
  geom_hline(yintercept = result@result$pwr_eff,
             linetype = 1,
             color = "red") +  
  geom_vline(xintercept = result@result$n_total,
             linetype = 1,
             color = "red") +  
  geom_point() +
  ggplot2::geom_errorbar(aes(
    ymin = pwr_eff - 1.96 * se_pwr_eff,
    ymax = pwr_eff + 1.96 * se_pwr_eff
  )) +
  geom_line() +
  scale_y_continuous(limits = c(0, 1)) +
  scale_x_continuous(limits = c(NA, NA)) +
  labs(title = "Sampled Points during Optimization", x = "Total Sample Size", y = "Power for Efficacy") +
  theme_minimal()
```


# Use Cases

## Alpha Control

```{r}
obj_alpha <- build_objectives(
 design = design,
 search = list(thr_dec_eff = c(.5,1)),
 objectives = list(pwr_eff = target(.05, direction = "<=")),
 constant = list(
   n_total = 200,
   b_arm_treat = 0,
   thr_dec_fut = 0.5,
   thr_fx_eff = 0.1,
   thr_fx_fut = 0,
   intercept = 0,
   b_covariate = 0.3,
   sigma = 1
 )
)
result_alpha <- optimization(
 obj_alpha,
 n_sims = 1000,
 max_evals = 500,
 surrogate = "rf",
 verbosity = 1,
 init_design_size = 4,
 refresh = 10,
 bf_args = list(
   n_posterior_samples = 1000
 )
)
```

```{r}
result_alpha@archive |>
  mutate(id_draw = row_number()) |>
  ggplot(aes(x = id_draw, y = pwr_eff)) +
  geom_hline(yintercept = 0.05,
             linetype = "dashed",
             color = "grey20") +
  geom_point() +
  ggplot2::geom_errorbar(aes(
    ymin = pwr_eff - 1.96 * se_pwr_eff,
    ymax = pwr_eff + 1.96 * se_pwr_eff
  )) +
  geom_line() +
  scale_y_continuous(limits = c(0, .5)) +
  labs(title = "Sampled Points during Optimization", x = "Draw Number", y = "Power for Efficacy") +
  theme_minimal()
result_alpha@archive |>
  mutate(id_draw = row_number()) |>
  ggplot(aes(x = thr_dec_eff, y = pwr_eff)) +
  geom_hline(yintercept = 0.05,
             linetype = "dashed",
             color = "grey20") +
  geom_point() +
  ggplot2::geom_errorbar(aes(
    ymin = pwr_eff - 1.96 * se_pwr_eff,
    ymax = pwr_eff + 1.96 * se_pwr_eff
  )) +
  geom_line() +
  scale_y_continuous(limits = c(0, .5)) +
  labs(title = "Sampled Points during Optimization", x = "Threshold for Efficacy Decision", y = "Power for Efficacy") +
  theme_minimal()
```


# Advanced Options

## Acquisition Functions

The acquisition function guides where to sample next. Use `acq_function` to select:

| Function | Description |
|----------|-------------|
| `"auto"` | Default: EI for single/target, scalarized EI for multi-objective |
| `"ei"` | Expected Improvement |
| `"cb"` | Confidence Bound (lower for minimization) |
| `"pi"` | Probability of Improvement |
| `"eic"` | Expected Improvement with Constraints (target optimization only) |

### EIC: Expected Improvement with Constraints

EIC is designed for **constrained optimization** where you want to find the minimum sample size that achieves a target power. It uses two surrogate models:

1. **Objective surrogate**: Predicts sample size (n_total)
2. **Constraint surrogate**: Predicts power on logit scale

The acquisition value is: `EIC(x) = EI(x) × P(feasible)`

where P(feasible) is the probability that power ≥ target, computed from the constraint surrogate's predictive distribution.

```{r eic-example}
#| eval: false

# EIC with neutral P(feasible) estimation
result <- optimization(
  obj,
  surrogate = "gp",
  acq_function = "eic",
  eic_kappa = 0  # Uses posterior mean for P(feasible)
)

# Conservative EIC: uses 97.5% lower confidence bound
result_conservative <- optimization(
  obj,
  surrogate = "gp",
  acq_function = "eic",
  eic_kappa = 1.96  # 97.5% LCB for P(feasible)
)
```

**Kappa values for conservatism:**

| `eic_kappa` | Interpretation |
|-------------|----------------|
| 0 | Neutral — uses posterior mean for P(feasible) |
| 1.96 | 97.5% lower confidence bound — more conservative |
| 2.58 | 99.5% LCB — very conservative |

Higher kappa means recommendations are more likely to achieve the target, but may suggest larger sample sizes.

## Penalty Functions

For target optimization, `penalty` controls how constraint violations are handled:

| Penalty | Formula | Use Case |
|---------|---------|----------|
| `"none"` | Quadratic cost below target | Default, balanced |
| `"hard"` | Large penalty when P(target) < α | Binary feasibility |
| `"soft"` | Proportional to violation | Smooth penalty |
| `"ratio"` | n / p_feas^κ | Risk-weighted cost |
| `"logit_distance"` | 1 - invlogit(\|logit(power) - logit(target)\|) | Symmetric distance |

### Logit-Distance: Symmetric Target Optimization

The logit-distance penalty provides a **symmetric cost function** that:

- Peaks at exactly the target (cost = 0.5)
- Penalizes undershoot and overshoot equally at the same logit distance
- Is smooth and differentiable everywhere

```{r logit-distance-example}
#| eval: false

result <- optimization(
  obj,
  surrogate = "gp",
  penalty = "logit_distance"
)
```

**Cost values for target = 0.80:**

| Power | Cost |
|-------|------|
| 0.80 | 0.500 (maximum) |
| 0.75 | 0.428 |
| 0.85 | 0.428 |
| 0.70 | 0.357 |
| 0.90 | 0.357 |
| 0.50 | 0.200 |

This is useful when you want to find designs that achieve **exactly** the target power, rather than exceeding it.

## Combining EIC with Penalties

EIC and penalty functions can be used together for robust constrained optimization:

```{r combined-example}
#| eval: false

result <- optimization(
  obj,
  surrogate = "gp",
  acq_function = "eic",
  eic_kappa = 1.96,           # Conservative feasibility
  penalty = "logit_distance"  # Symmetric target matching
)
```

***

# Session Information

```{r session-info}
sessionInfo()
```
