---
title: "Bayesian Optimization for Trial Design"
author:
 - name: Matthias Kloft
   orcid: 0000-0003-1845-6957
date: "`r Sys.Date()`"
format:
  html:
    toc: true
    number-sections: true
    theme: cosmo
    code-fold: false
    code-tools: true
    code-summary: "Show the code"
    fig-width: 7
    fig-height: 5
    embed-resources: true
execute:
  message: true
  warning: true
---

```{r setup}
#| message: false

library(rctbayespower)
library(tidyverse)


backend <- "bf"

if(backend == "bf") {
  # Initialize BayesFlow Python environment
  init_bf_python("rctbp-3-12-gpu")
}

```

***

# Introduction

Finding optimal trial designs often requires exploring a large parameter space. Traditional grid search evaluates every combination, which becomes computationally prohibitive as dimensions grow.

**Bayesian Optimization (BO)** efficiently searches continuous parameter spaces by:

1. Building a surrogate model (Gaussian Process) of the objective function
2. Using an acquisition function to select promising points to evaluate
3. Iteratively refining the search based on observed results

This is ideal for power analysis where each evaluation (running simulations) is expensive.

## Grid Search vs. Bayesian Optimization

| | Grid Search | Bayesian Optimization |
|--------|-----------|-----------|
| **API** | `build_conditions()` → `power_analysis()` | `build_objectives()` → `optimization()` |
| **Search** | Discrete grid points | Continuous bounds |
| **Evaluations** | All points | Guided selection |
| **Use case** | Visualization, known ranges | Sample size determination |

***

# Workflow

## Step 1: Create Design

Use the BayesFlow backend for faster evaluations.
```{r create-design}
design <- build_design(
 model_name = "ancova_cont_2arms",
 backend = backend,
 target_params = "b_group"
)
```

## Step 2: Define Optimization Problem

Use `build_objectives()` to specify:

- **search**: Parameter bounds to optimize over
- **objectives**: What to achieve (target power, minimize sample size)
- **constant**: Fixed design parameters

```{r}
show_objective_fns()
#obj_min_n_for_power()
```


```{r build-objectives}
obj <- build_objectives(
 design = design,
 search = list(n_total = c(20, 1000)),
 objectives = list(pwr_eff = target(0.9, direction = ">=")),
 constant = list(
   p_alloc = .5,
   b_arm_treat = 0.3,
   thr_dec_eff = 0.95,
   thr_dec_fut = 0.5,
   thr_fx_eff = 0.1,
   thr_fx_fut = 0,
   intercept = 0,
   b_covariate = 0.3,
   sigma = 1
 )
)
#print(obj)
```

## Step 3: Run Optimization


### Configuration Options

```{r show-args}
#show_optimization_args()
```

### Run Optimization
```{r run-optimization, results='hide'}
#| eval: false

obj <- build_objectives(
 design = design,
 search = list(n_total = c(20, 1000)),
 objectives = list(pwr_eff = target(0.9, direction = ">=")),
 constant = list(
   p_alloc = .5,
   b_arm_treat = 0.3,
   thr_dec_eff = 0.95,
   thr_dec_fut = 0.5,
   thr_fx_eff = 0.1,
   thr_fx_fut = 0,
   intercept = 0,
   b_covariate = 0.3,
   sigma = 1
 )
)
result <- optimization(
 obj,
 n_sims = c(100, 101, 500),
 evals_per_step = c(20, 20, 20),
 use_warmup = TRUE,
 min_delta = .01,  # Minimum improvement
 patience = 10,
 surrogate = "rf",   # Gaussian Process surrogate
 verbosity = 1, 
 init_design_size = 5,
 refresh = 10,
 bf_args = list(
   n_posterior_samples = 1000
 )
)
```

```{r}
result@optimization_type
```

### Plot sampled points: n_total vs. time

```{r}
result@archive |>
  mutate(id_draw = row_number()) |>
  ggplot(aes(x = id_draw, y = pwr_eff)) +
  geom_hline(yintercept = 0.9,
             linetype = "dashed",
             color = "grey20") +
  geom_hline(yintercept = result@result$pwr_eff,
             linetype = 1,
             color = "red") +           
  geom_point() +
  ggplot2::geom_errorbar(aes(
    ymin = pwr_eff - 1.96 * se_pwr_eff,
    ymax = pwr_eff + 1.96 * se_pwr_eff
  )) +
  geom_line() +
  scale_y_continuous(limits = c(0, 1)) +
  labs(title = "Sampled Points during Optimization", x = "Draw Number", y = "Power for Efficacy") +
  theme_minimal()

result@archive |>
  mutate(id_draw = row_number()) |>
  ggplot(aes(x = n_total, y = pwr_eff)) +
  geom_hline(yintercept = 0.9,
             linetype = "dashed",
             color = "grey20") +
  geom_hline(yintercept = result@result$pwr_eff,
             linetype = 1,
             color = "red") +  
  geom_vline(xintercept = result@result$n_total,
             linetype = 1,
             color = "red") +  
  geom_point() +
  ggplot2::geom_errorbar(aes(
    ymin = pwr_eff - 1.96 * se_pwr_eff,
    ymax = pwr_eff + 1.96 * se_pwr_eff
  )) +
  geom_line() +
  scale_y_continuous(limits = c(0, 1)) +
  scale_x_continuous(limits = c(200, NA)) +
  labs(title = "Sampled Points during Optimization", x = "Total Sample Size", y = "Power for Efficacy") +
  theme_minimal()
```


# Use Cases

## Alpha Control

```{r}
obj_alpha <- build_objectives(
 design = design,
 search = list(thr_dec_eff = c(.5,1)),
 objectives = list(pwr_eff = target(.05, direction = "<=")),
 constant = list(
   n_total = 200,
   b_arm_treat = 0,
   thr_dec_fut = 0.5,
   thr_fx_eff = 0.1,
   thr_fx_fut = 0,
   intercept = 0,
   b_covariate = 0.3,
   sigma = 1
 )
)
result_alpha <- optimization(
 obj_alpha,
 n_sims = 1000,
 max_evals = 500,
 surrogate = "rf",
 verbosity = 1,
 init_design_size = 4,
 refresh = 10,
 bf_args = list(
   n_posterior_samples = 1000
 )
)
```

```{r}
result_alpha@archive |>
  mutate(id_draw = row_number()) |>
  ggplot(aes(x = id_draw, y = pwr_eff)) +
  geom_hline(yintercept = 0.05,
             linetype = "dashed",
             color = "grey20") +
  geom_point() +
  ggplot2::geom_errorbar(aes(
    ymin = pwr_eff - 1.96 * se_pwr_eff,
    ymax = pwr_eff + 1.96 * se_pwr_eff
  )) +
  geom_line() +
  scale_y_continuous(limits = c(0, .5)) +
  labs(title = "Sampled Points during Optimization", x = "Draw Number", y = "Power for Efficacy") +
  theme_minimal()
result_alpha@archive |>
  mutate(id_draw = row_number()) |>
  ggplot(aes(x = thr_dec_eff, y = pwr_eff)) +
  geom_hline(yintercept = 0.05,
             linetype = "dashed",
             color = "grey20") +
  geom_point() +
  ggplot2::geom_errorbar(aes(
    ymin = pwr_eff - 1.96 * se_pwr_eff,
    ymax = pwr_eff + 1.96 * se_pwr_eff
  )) +
  geom_line() +
  scale_y_continuous(limits = c(0, .5)) +
  labs(title = "Sampled Points during Optimization", x = "Threshold for Efficacy Decision", y = "Power for Efficacy") +
  theme_minimal()
```


***

# Objective Types

## Target Constraint

Find minimum sample size achieving target power:

```{r target-example}
#| eval: false

obj <- build_objectives(
 design = design,
 search = list(n_total = c(50, 500)),
 objectives = list(pwr_eff = target(0.80)),
 constant = list(...)
)
```

## Multi-Objective (Pareto)

Explore power vs. sample size tradeoff:

```{r multi-objective}
#| eval: false

obj <- build_objectives(
 design = design,
 search = list(
   n_total = c(50, 500),
   b_arm_treat = c(0.2, 0.6)
 ),
 objectives = list(
   pwr_eff = "maximize",
   n_total = "minimize"
 ),
 constant = list(...)
)

result <- optimization(obj, n_sims = 200, n_evals = 50)
plot(result, type = "pareto")
```

***

# Visualization

```{r plot-example}
#| eval: false

# Convergence trace
plot(result, type = "convergence")

# Search space exploration
plot(result, type = "search")

# Pareto front (multi-objective only)
plot(result, type = "pareto")
```

***



## Key Parameters

- **n_sims**: More simulations = less noise, but slower. Start with 100-200.
- **n_evals**: Budget for objective evaluations. 20-50 typical.
- **surrogate**: `"gp"` (Gaussian Process) or `"rf"` (Random Forest)
- **initial_design_size**: Initial points before BO starts. Default: 4 × dimensions.

***

# Session Information

```{r session-info}
sessionInfo()
```
