---
title: "Getting Started with rctbayespower: Bayesian Power Analysis for RCTs"
date: "`r Sys.Date()`"
always_allow_html: true
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "vignettes/man/figures/04-algorithm_performance-",
  out.width = "90%",
  fig.width = 10,
  fig.height = 7,
  message = FALSE
)

# Check and install required packages
packages <- c("tidyverse", "devtools")
if (!require("pacman")) install.packages("pacman") 
pacman::p_load(packages, update = F, character.only = T)

# Load the package
devtools::load_all(".")

# Fast performance testing settings
n_cores <- parallel::detectCores() - 1
  n_sims <- 1e3
```



# Algorithm Performance and Computational Considerations

This vignette provides guidance on algorithm selection, performance optimization, and computational considerations when using the `rctbayespower` package for Bayesian power analysis.


## Algorithm Overview

The `rctbayespower` package supports multiple algorithms with different speed-accuracy trade-offs:

**Recommended workflow:**
1. **Exploration**: `meanfield` for faster roundtrip
2. **Refinement**: `fullrank` as an alternative to `meanfield` for more accuracy, however might not converge
3. **Confirmation**: `sampling` for final results



# Performance Comparison

```{r basic_performance}
model_compiled <- power_analysis_ancova(
  n_control = 100,
  n_treatment = 100,
  effect_size = 0.6,
  baseline_effect = 0.3,
  outcome_type = "continuous",
  compile_models_only = TRUE
)

# Compare core algorithms on standard problem
performance_results <- data.frame(
  Algorithm = character(),
  Time_Seconds = numeric(),
  Power_Success = numeric(),
  Probability_Success = numeric(),
  Power_Futility = numeric(),
  Probability_Futility = numeric(),
  Convergence = numeric(),
  stringsAsFactors = FALSE
)

algorithms <- c("meanfield", "fullrank", "sampling")

for (alg in algorithms) {
  cat("Testing algorithm:", alg, "\n")
  
  start_time <- Sys.time()
  
  # Use power_analysis_ancova for meanfield and fullrank
  if (alg %in% c("meanfield", "fullrank")) {
    power_result <- power_analysis(
      brms_design_estimation = model_compiled$brms_design_estimation,
      brms_design_true_params = model_compiled$brms_design_true_params,
      simulate_data_fn = model_compiled$simulate_data,
      target_param = model_compiled$target_param,
      threshold_success = .5,
      threshold_futility = .1,
      p_sig_success = .975,
      p_sig_futility = .5, 
      n_simulations = n_sims,
      n_cores = n_cores,
      brms_args = list(
        algorithm = alg,
        importance_resampling = TRUE,
        iter = 1e4,
        output_samples = 1e3
      )
    )
  }
  # Use power_analysis for sampling
  if (alg == "sampling") {
    power_result <- power_analysis(
      brms_design_estimation = model_compiled$brms_design_estimation,
      brms_design_true_params = model_compiled$brms_design_true_params,
      simulate_data_fn = model_compiled$simulate_data,
      target_param = model_compiled$target_param,
      threshold_success = .5,
      threshold_futility = .1,
      p_sig_success = .975,
      p_sig_futility = .5, 
      n_simulations = n_sims,
      n_cores = n_cores,
      brms_args = list(
        chains = 4,
        iter = 800,
        warmup = 300
      )
    )
  }
  
  elapsed_time <- as.numeric(difftime(Sys.time(), start_time, units = "secs"))
  # time in hours:minutes:seconds
  elapsed_time_formatted <- sprintf(
    "%02d:%02d:%02d",
    as.integer(elapsed_time / 3600),
    as.integer((elapsed_time %% 3600) / 60),
    as.integer(elapsed_time %% 60)
  )
  
  performance_results <- rbind(
    performance_results,
    data.frame(
      Algorithm = alg,
      N_simulations = n_sims,
      Time = elapsed_time_formatted,
      Power_Success = power_result$power_success,
      Probability_Success = power_result$mean_prob_success,
      Power_Futility = power_result$power_futility,
      Probability_Futility = power_result$mean_prob_futility,
      Convergence = power_result$convergence
    )
  )
}

print(performance_results) |> 
  kableExtra::kable(digits = 2
  ) 
```


# Conclusion

- **Start fast**: Use `meanfield` for exploration
- **Refine carefully**: Use `fullrank` for detailed analysis  
- **Confirm precisely**: Use `sampling` for final results
- **Plan resources**: Estimate time and memory needs
- **Monitor quality**: Check convergence and validate results



***


# Session Information

```{r session_info}
sessionInfo()
```
